A Java Wrapper For Embarrassingly Parallel Problems

### 2013-11-21 paper/paper.tex
Wills paper revisions

\begin{abstract}
An embarrassingly parallel problem, also called pleasingly parallel, is one that 
can easily be broken up into identital components that do not need to interact 
to find the solution. Several problems in the biological sciences are 
embarrsingly parallel but they are difficult to rewrite to use multithreading, 
because they are either poorly written or too complex. We have written a 
framework, implemented in Java, that will execute the serial program in several 
threads. These threads working together help cut down the total runtime
exponentially. The next step is to seperate the framework into a front-end that
will run locally and the back-end which recieves blocks of work and does the
computations. This back-end should be running in the cloud, on a machine who is
dedicated to running embarrassingly parallel tasks only.
\end{abstract}

@@

A computation graph consists of nodes that represent the functions or processes 
and the data being passed from one process to another is displayed in the graph 
as edges. If the graph is disconnected, none of the processes need to 
communicate to complete the computations. Such a computation is called
"embarrassingly parallel." In a practical situation, there will be some 
internode communication to set up the problem and then again to accumulate the 
results back together. Also known as pleasingly parallel problems, they can 
easily be performed on a cluster of servers or in any kind of distributed 
network, due to its lack of dependencies. The term embarrassingly parallel was 
coined by Cleve Moler in his 1986 paper "Matrix Computation on Distributed
Memory Multiprocessors." \cite{history}

Programs in the field of Biology and Genetics often analyze long strings of
characters, whether they are DNA, RNA or protein sequences. In addition to the
large file sizes, there is a lot of processing that goes into comparing and
manipulating strings. This time adds up when there are thousands of strings to 
process. However, these sequences often do not relate to each other and can be 
processed independently. Using an embarrassingly parallel approach, the 
processing time for all of the sequences can often be cut down exponentially 
with very minimal effort by the programmer.

@@

The motivation for this effort was with research into different forms of the 
species \emph{Cicer arietinum}, the chickpea. The two main forms, Kabuli and 
Desi, are very different in size and color but we mostly wanted to see why the 
Desi is more disease resistance and have more effective root nodule symbiosis 
and why the Kabuli has a large but soft shell. These two types are both 
cultivated and have been under human selection for several thousand years. We 
wanted to use a Ka/Ks equation to quantify the evolution and selection of these 
chickpeas. We found a Ka/Ks program writen by Zhang et al. \cite{kaks} but had 
three variants that we wanted to be analyzed, each with thousands of sequences. 
The Ka/Ks Calculator program takes somewhere from five to thirty seconds to 
process one pair of sequences.

Instead of attempting to rewrite the preexisting program, we decided it would be
better to look at other options to process the sequences faster. Trying to 
parallelize the serial Ka/Ks program could present several problems: introducing
bugs, breaking the algorithm, dealing with the learning curve of C++ threading. 
We decided a  different approach was neccessary, mostly due to the short time we
had to produce results. A simple Java program was written to break up the input 
and spawn  threads that would run the Ka/Ks Calculator process. When the input 
was calculated, the contents of the output files was concatenated together to 
produce a final output that mnatches the format and order that would have been
produces by Ka/Ks Calculator.

@@

\section{Proposed Solution}
\begin{figure}
{\resizebox{5.2in}{!}{\includegraphics{figures/uml.png}}}
\caption{UML for the embarrassingly parallel framework}
\label{fig:uml}
\end{figure}
The only requirements for a program to be wrapped by framework are that it 
process the input in an embarrassingly parallel manner and that the input comes 
from a file. The processed results are must also written to a file. The 
framework supports two forms of input: a directory of files that contain that 
each contain one unit of work, which is represented by the Chunk class, or a 
single file that the framework can break up into Chunks. For example, the Ka/Ks 
Calculator's input consists of a header and then two protein sequence lines 
followed by a blank line. The framework uses that blank line to know when one 
Chunk of work has ended and another has begun. The user can input a regular 
expression to define when lines of input are seperated into Chunks. This occurs 
in the ChunkManager class.

The other purpose of the ChunkManager class is to combine the resulting Chunks 
back together when the worker threads have finished. Once the work has been
done, all of the separate results are spread across several files. The
ChunkManager's second job is to retrieve those results and construct a real
output file. The real output file must look the same as the output file would
look if the program were run serially; the output must come in the same order as
the inputs in the input file and headers may need to be deleted from the
separate files and added to the combined file. There are two other options if
the end-user does not want this behavior. They can choose to write their own
merger in Java, using the CustomMerge class, or write their own script to do the
merging.

@@

To observe how the framework changes the runtime of sequential programs when
using varying numbers of threads, we designed a simple experiment. The Ka/Ks 
Calculator was run through the framework with ten different samples, each 
containing sixteen units of work. Each sample was run by the framework at five 
different level of numbers of worker threads, starting at 1 and increasing by 
powers of two until the 100\% CPU usage (16 processors). These tests were run
with minimal other activity on the server.

@@

As the numbers of threads increased, the time it took descreased exponentially. 
In a few cases, there was an increase in time when attempting to max out the CPU 
usage. This occured because the framework had to wait until for the CPU to be 
finished with its previous work before it was allowed to start work on what the 
framework wanted it to do. In the cases when the framework executed faster while
the CPU was maxed out, it was only faster a minute at the most. With such a
small time gain with the CPU at 50\% versus the CPU at 100\%, running the
framework at 50\% CPU is more efficient when the number of threads is on the
same order as the number of units of work.

%TODO: Rewrite with new next steps
There are several next steps in the pipeline. In addition to using the
statistics to try to boost speed, adding savable configurations is at the top of
the list. This will allow users to use the same configuration without having to
go through the wizard if they need to redo a run. More extensive unit testing
will be implemented. The ability to chunk multiple pieces of work may be
implemented in order to avoid potential start up costs in the executable. 
Finally, other pleasing parallel problems will be executed by the wrapper so 
that more potential features will be discovered.

%TODO: Does this still make sense
Using Java was a good choice for implementation ease and future implementation
extension by others, but it may not produce the fastest results. Languages
like Hadoop, Clojure or other multithread oriented language may have been better
due to their parallel nature. In addition to a single machine working to process
the data quickly, we can also look to the cloud for help by using a networked 
solution like MPI or a cloud solution like Amazon instances. This would
potentially allow for an even greater time decrease because the number of
machines that could help is most likely larger than the number of processors on
one machine.

>>>>

\begin{abstract}
An embarrassingly parallel problem, also called pleasingly parallel, is one that 
can be easily broken up into identital components that do not need to interact 
to find the solution. Several problems in the biological sciences are 
embarrsingly parallel but they are difficult to rewrite to use multithreading, 
because they are either poorly written or too complex. We have written a 
framework, implemented in Java, that will execute the serial program in several 
threads. These threads working together help cut down the total runtime of the
program exponentially. The next step is to seperate the framework into a 
front-end that will run locally and a back-end which recieves blocks of work and 
does the computations. This back-end should run in the cloud, on a machine that 
is dedicated to running embarrassingly parallel tasks only.
\end{abstract}

@@

A computation graph consists of nodes that represent the functions or processes. 
The edges of the graph show data being passed from one process to another. If 
the graph is disconnected, none of the processes need to communicate to complete
the computations. Such a computation is called "embarrassingly parallel." In a 
practical situation, there will be some internode communication to set up the 
problem, and then again to accumulate the results. Embarrassingly parallel 
problems can be easily performed on a cluster of servers or in any kind of 
distributed network, due to its lack of dependencies. The term embarrassingly 
parallel was coined by Cleve Moler in his 1986 paper "Matrix Computation on 
Distributed Memory Multiprocessors." \cite{history}

Programs in the field of Biology and Genetics often analyze long strings of
characters, whether they are DNA, RNA or protein sequences. In addition to
handling large file sizes, there is a lot of processing that goes into comparing
and manipulating strings. This time adds up when there are thousands of strings 
to process. However, these sequences often do not relate to each other and can 
be processed independently. Using an embarrassingly parallel approach, the 
processing time for all of the sequences can often be cut down exponentially 
with very minimal effort by the programmer.

@@

The motivation for this effort was research into different forms of the 
species \emph{Cicer arietinum}, the chickpea. The two main forms, Kabuli and 
Desi, are very different in size and color. The Desi is also more resistant to 
diseases and are more effective at root nodule symbiosis. The Kabuli has a large 
but soft shell, which make it at better food source for humans. These two types 
are both cultivated and have been under human selection for several thousand 
years. We wanted to use a Ka/Ks equation to quantify the evolution and selection 
of these chickpeas. We found a Ka/Ks program written by Zhang et al. \cite{kaks}
but had three variants that we wanted to analyze, each with thousands of 
sequences. The Ka/Ks Calculator program takes somewhere from five to thirty 
seconds to process one pair of sequences.

Instead of attempting to rewrite the preexisting program, we decided it would be
better to look at other options to process the sequences faster. Trying to 
parallelize the serial Ka/Ks program could present several problems: introducing
bugs, breaking the algorithm, dealing with the learning curve of C++ threading. 
We decided that a  different approach was neccessary, mostly due to the short 
time given to produce results. A simple Java program was written to break up the
input and spawn threads that would run the Ka/Ks Calculator process. When the 
input was calculated, the contents of the output files were concatenated 
together to produce a final output that matches the format and order that would 
have been produced by the Ka/Ks Calculator program.

@@

\section{Proposed Solution}
\begin{figure}
{\resizebox{5.2in}{!}{\includegraphics{figures/uml.png}}}
\caption{UML for the embarrassingly parallel framework}
\label{fig:uml}
\end{figure}
The only requirements for a program to be wrapped by framework are that it 
process the input in an embarrassingly parallel manner and that the input comes 
from a file. The processed results must also written to a file. The framework 
supports two forms of input: a directory of files that contain that each contain
one unit of work, which is represented by the Chunk class, or a single file that
the framework can break up into Chunks. For example, the Ka/Ks Calculator's 
input consists of a header and then two protein sequence lines followed by a 
blank line. The framework uses that blank line to know when one Chunk of work 
has ended and another has begun. The user can input a regular expression to 
define when lines of input are seperated into Chunks. This occurs in the 
ChunkManager class.

The other purpose of the ChunkManager class is to combine the resulting Chunks 
back together when the worker threads have finished. Once the work has been
done, all of the separate results are spread across several files. Those results 
must be retrieved and combined to construct the final output file. This output 
file must look the same as the output file would look if the program were run 
serially; the output must come in the same order as the inputs in the input file
and headers may need to be deleted from the separate files and added to the 
combined file. There are two other options if the end-user does not want this 
behavior. They can choose to write their own merger in Java, using the 
CustomMerge class, or write their own script to do the merging.

@@

In order to observe how the framework changes the runtime of sequential programs 
when using varying numbers of threads, we designed a simple experiment. The 
Ka/Ks Calculator was run through the framework with ten different samples, each 
containing sixteen units of work. Each sample was run by the framework at five 
different level of numbers of worker threads, starting at 1 and increasing by 
powers of two until the 100\% CPU usage (16 processors). These tests were run
with minimal other activity on the server.

@@

As the numbers of threads increased, the runtime of the program descreased 
exponentially. In a few cases, there was an increase in time when attempting to 
max out the CPU usage. This occured because the framework had to wait for the 
CPU to be finished with its previous work before it was allowed to start any 
further jobs. In the cases when the framework executed faster while the CPU was 
maxed out, it was only faster a minute at the most. However, running the 
framework at 50\% of the CPU is more efficient than running using 100\% of the 
CPU. The runtime drop achieved when using 100\% of the CPU versus 50\% is not 
large enough to warrant using all of the machine's power.

Using Java was a good choice for implementation ease and future implementation
extension by others, but it may not produce the fastest results. Languages
like Hadoop, Clojure or other multithread oriented language may have been a
better choice due to their parallel nature. In addition to a single machine 
working to process the data quickly, we can also look to the cloud for help by 
using a networked solution like MPI or a cloud solution like Amazon instances. 
In the future, we would like to see this work be split into a local front-end
and a cloud back-end. This would potentially allow for an even greater time 
decrease because the number of machines that could help is much larger than the 
number of processors on one machine. Spliting the framework would also allow for
better testing of cloud environments, where the cloud environment itself could
be a configurable option for end users.
